---
title: "Project2"
author: "Christian Lehre, Erik Bøe & Axel Ronold"
date: "3/5/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = "hold")
knitr::opts_chunk$set(error=FALSE)
knitr::opts_chunk$set(warning=FALSE)
```

# Problem 1

**a)**
Testing Merge
    
* `Estimate` - in particular interpretation of `Intercept`
\newline 
Estimated regression coefficients given by 
$$\hat{\boldsymbol{\beta}} = (X^TX)^{-1}X^T\mathbf{Y}$$.
\newline When increasing covariate $x_j$ with one unit, and keeping all other covariates constant, the response variable changes with a factor of $\hat{\beta_j}$. Positive estimates reduce the value of the response, while negative estimates increase the value. Example increasing AGE from 25 to 26 will increase $\log(\text{FEV})$ by $0.023$. Similarily, for the binary predictor SMOKE, the coefficient estimate represent the change in the response when changing from non-smoker (0) to smoker (1), and is equal to $-0.046$. The intercept is the value of the response when all covariates are set to zero. This is not necessarily realistic, as there is no humans with e.g zero height.

* `Std.Error`
\newline 
Estimated standard deviation of the estimated regression coefficients, i.e the average amount that the estimated regression coefficients vary from the actual value. The Std.Error is given by 
$$\widehat{\text{SD}}(\hat{\beta_j}) =\sqrt{\hat{\sigma}^2(X^TX)^{-1}_{jj}}$$, where $X$ is the design matrix of the regression, and $\hat{\sigma}$ is the residual standard error.

* `t value`

*`P(>|t|)`


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
